Local Spectral Clustering of Density Upper Level Sets

### 2019-01-23 icml/icml_experiments.tex
final icml version

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}
\newcommand{\theHalgorithm}{\arabic{algorithm}}

%%%%%
\icmltitlerunning{Local Spectral Clustering of Density Upper Level Sets}

@@

\twocolumn[
\icmltitle{Local Spectral Clustering of Density Upper Level Sets}
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Alden Green}{cmu}
\icmlauthor{Sivaraman Balakrishnan}{cmu}
\icmlauthor{Ryan J. Tibshirani}{cmu}
\end{icmlauthorlist}

\icmlaffiliation{cmu}{Department of Statistics and Data Science, Carnegie Mellon
  University, Pittsburgh PA, USA} 
\icmlcorrespondingauthor{Alden Green}{ajgreen@andrew.cmu.edu}
\icmlkeywords{local clustering}

@@

\begin{abstract}
\vskip 0.1 in % RJT: there was no space here and it looked very squished 
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several authors have shown that local methods, such as personalized
PageRank (PPR), have worst-case guarantees for certain graph-based measures of
cluster quality.  In contrast to existing works, we analyze PPR in a traditional
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.
\end{abstract}

\section{Introduction}
\label{sec: introduction}

Let $\mathbf{X} = \{x_1, \ldots, x_n\}$ be a sample drawn i.i.d.\ from a
distribution $\Pbb$ on $\Rd$, with density $f$, and consider the problem of 
clustering: splitting the data into groups which satisfy some notion of
within-group similarity and between-group difference.  We focus on spectral
clustering methods, a family of powerful nonparametric clustering algorithms.
Roughly speaking, a spectral technique first constructs a geometric graph $G$,
where vertices are associated with samples, and edges correspond to proximities
between samples. It then learns a feature embedding based on the Laplacian of
$G$, and applies a simple clustering technique (such as k-means clustering) in
the embedded feature space.

To be more precise, let $G=(V,E,w)$ denote a weighted, undirected graph  
constructed from the samples $\mathbf{X}$, where $V=\{1,\ldots,n\}$, and $w_{uv}
= K(x_u,x_v) \geq 0$ for $u,v \in V$, and a particular kernel function $K$.
Here $(u,v) \in E$ if and only if $w_{uv} > 0$.  We denote by $\Abf \in
\Reals^{n \times n}$ the weighted adjacency matrix, which has entries
$A_{uv}=w_{uv}$, and by $\Dbf$ the degree matrix, with 
\smash{$\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$}.  We also denote by $\Wbf,\Lbf$
the random walk transition probability matrix and normalized\footnote{Other
  popular choices here include the unnormalized Laplacian, and symmetric
  normalized Laplacian.} 
Laplacian matrix, respectively, which are defined as
$$
\Wbf = \Dbf^{-1}\Abf, \quad \Lbf = \Ibf{} - \Wbf,
$$
where $\Ibf{} \in \Reals^{n\times n}$ is the identity matrix.  Classical global
spectral methods take a eigendecomposition $L=U \Sigma U^T$, use some 
number of eigenvectors (columns in $U$) as a feature representation for the
samples, and then run (say) k-means in this new feature space.

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be been computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the PPR
algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citealt{hartigan1981}).

\subsection{Graph Connectivity Criteria}

Here we define a pair of criteria that reflect the quality of a cluster with
respect to $G=(V,E,w)$.  There are many graph-based measures of cluster quality  
that one could consider; see, e.g., \citet{yang2015,fortunato2010} for 
an overview.  The pair of criteria that we focus on are (arguably) quite
natural, and moreover, they play a fundamental role in our analysis 
of the PPR algorithm.  Our two critera capture the \emph{external} and
\emph{internal} connectivity of a subset $S \subseteq V$, denoted $\Phi(S; G)$
and $\Psi(S; G)$, respectively, and defined below in turn.

\paragraph{External Connectivity: Normalized Cut}

Define the cut between subsets $S,S' \subseteq V$ to be
$$
\cut(S, S'; G) = \sum_{u \in S} \sum_{v \in S'} w_{uv},
$$
and define
\smash{$\vol(S; G) = \cut(S, V; G) = \sum_{u \in S} \sum_{v \in V} w_{uv}$}.  
As our notion of external connectivity, we use the \emph{normalized cut}  
of $S$, defined as 
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) = \frac{\cut(S; G)}{\min\{\vol(S; G), \vol(S^c; G)\}},
\end{equation} 
where we abbreviate $\cut(S; G) = \cut(S; S^c; G)$.

\paragraph{Internal Connectivity: Inverse Mixing Time}

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
$$
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the rotation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\widetilde\pibf = (\widetilde\pi_{u})_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that the stationariy distribution is given by
\smash{$\widetilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ t: \frac{|q_{vu}^{(m)} - 
      \widetilde\pi_u|}{\widetilde\pi_u} \leq \frac{1}{4}, 
    \; \text{for $u,v \in V$}},  
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no
less than $\Psi$, we call it as a \emph{$(\Phi,\Psi)$-cluster}. Both
local \citep{zhu2013} and global \citep{kannan04} spectral algorithms have been
shown to output clusters (or partitions) which approximate the optimal $(\Phi,
\Psi)$-cluster (or partition) for a given graph $G$.\footnote{In the case of
  \citet{kannan04}, the internal connectivity parameter $\phi$ is actually the
  conductance, i.e., the minimum normalized cut within the subgraph $G[S]$. See
  Theorem 3.1 in their paper for details; however, note that $\phi^2 /
  \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a
  lower bound on $\Psi$.}

\subsection{PPR on a Neighborhood Graph}

We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted graph with vertices
$V=\{1,\ldots,n\}$, and an edge $(u,v) \in E$ if and only if $\norm{x_u - x_v}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm.  Note that this is a
special case of the general construction introduced above, with 
$K(u,v) = 1(\norm{x_u - x_v} \leq r)$.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, with respect to  
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf$ is the random walk matrix of the underlying graph $G_{n,r}$ 
and $e_{v}$ denotes indicator vector for node $v$ (with a 1 in the $v$th
position and 0 elsewhere).  In practice, we can approximately solve the above
linear system via a simple, efficient random walk, with appropriate restarts to
$v$.

For a level $\beta > 0$ and a target volume $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of PPR
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target volume $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return \smash{$\widehat{C} = S_{\beta^*}$}, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Summary of Results}

% It is worth calling attention to some other work on computing the normalized
% cut over neighborhood graphs. In this context, continuous analogues to (for
% instance) normalized cut have been defined, over the data-manifold rather than
% the graph, and convergence finite sample graph-theoretic functionals to their
% continuous counterparts has been shown 
% \cite{garciatrillos16, arias-castro12, maier11}. 
% However, in addition to the graph-minimization problem being computationally
% infeasible, these continuous analogues are not always easily interpretable --
% and their corresponding minimizers not always easily identifiable -- for the
% particular density function under consideration. Of course, relating these
% partitions to the arguably more simply defined high density clusters can be
% also challenging in general. Intuitively, however, under the right conditions
% such high-density clusters should have more edges within themselves than to
% the remainder of the graph. We formalize this intuition next.

%% RJT: I didn't know where to put this.  It was out of place, and now I'm not
%% sure where it goes ... should it go in related work?

%It is worth pointing out that in this context, some theory has been developed
%regarding how graph theoretic quantities such as the normalized cut $\Phi$ (and
%others) relate to properties of the underlying distribution $f$ as well as the
%kernel function $k$. Such analyses typically proceed by defining a continuous
%analogue to the measure of cluster quality under consideration. Then, under
%appropriate specification of $k$ and a proper schedule of $(r_n)$, convergence
%of clusters output by spectral (and other) algorithms to the corresponding
%minima of these continuous analogues has been shown \cite{vonluxburg2008,
%garciatrillos18}.

%% RJT: This was already commented out before

Let $\Cbb_f(\lambda)$ denote the connected components of the density upper level
set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster $\Cset \in
\Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical
  density cluster}. Below we define a notion of consistency in density cluster
estimation.

\begin{definition}[Consistent density cluster estimation]
\label{def: consistent_density_cluster_estimation}
For an estimator \smash{$\widehat{\Cset} \subseteq \Xbf$} and cluster 
$\Cset \in \Cbb_f(\lambda)$, we say \smash{$\widehat{\Cset}$} is a consistent
estimator of $\Cset$ if for all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset \not=
\Cset'$ the following holds as $n \to \infty$: 
\begin{equation}
\label{eqn: consistent_density_cluster_recovery}
\Cset[\Xbf] \subseteq \widehat{\Cset} \quad \text{and} \quad
\widehat{\Cset} \cap \Cset'[\Xbf] = \emptyset,
\end{equation}
with probability tending to 1.
\end{definition}

A summary of our main results (and outline for the rest of this paper) is as
follows.

\begin{enumerate}
\item In Section \ref{sec: cluster_quality_bounds}, we derive in Theorem
  \ref{thm: conductance_upper_bound} an upper bound on the normalized cut of a  
  (thickened) empirical density cluster $\Csig[\Xbf]$, under natural geometric 
  conditions (precluding clusters that are too thin and long).

\item Under largely the same set of geometric conditions, we derive in Theorem
  \ref{thm: inverse_mixing_time_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Csig[\Xbf]$.  We also provide in 
  Theorem \ref{thm: inverse_mixing_time_lower_bound} a tighter lower bound, but
  under more restrictive assumptions (convexity of $\Csig$).

\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show in
  Theorem \ref{thm: consistent_recovery_of_density_clusters} that these bounds
  in bounds in Theorems \ref{thm: conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex},
  on the cluster quality criteria, have algorithmic consequences for PPR:
  properly initialized, Algorithm \ref{alg: ppr} performs consistent density
  cluster estimation in the sense of \eqref{eqn:
    consistent_density_cluster_recovery}.

\item We show in Corollary \ref{cor: ppr_cluster} that Theorems \ref{thm:
    conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex}, along with the results in 
  \citet{zhu2013}, lead to alternative, graph-theoretic guarantees on cluster 
  quality: an upper bound on the normalized cut of the estimated cluster 
  \smash{$\widehat{C}$}, and an upper bound on volume of the symmetric set
  difference between \smash{$\widehat{C}$} and $\Cset[\Xbf]$.

\item In Section \ref{section: examples}, we discuss the implications of our
  results for some example density functions of interest, and empirically
  demonstrate that violations of the gometric conditions we require manifestly
  impact density cluster recovery. 
\end{enumerate}

On the topic of conditions, it is worth mentioning that, as density clusters
are inherently local, focusing on the PPR algorithm actually eases our analysis
and allows us to require fewer global regularity conditions relative to those
needed for more classical global spectral algorithms.

\subsection{Related Work}

In addition to the background given above, a few related lines of work are worth
highlighting. Global spectral clustering methods were first developed in the
context of graph partitioning \citep{fiedler1973,donath1973} and their
performance is well-understood in this context (see, e.g.,
\citealt{tolliver2006,luxburg2007}).  In a similar vein, several recent works
\citep{mcsherry2001,rohe2011,kamalika2012,balakrishnan2011,lei2015,abbe2018} 
have studied the efficacy of spectral methods in successfully recovering the
community structure in the stochastic block model and variants.

\citet{vonluxburg2008,hein2005}, building on earlier work of
\citet{koltchinskii2000}, studied the limiting behaviour of spectral clustering
algorithms. These authors show that when samples are obtained from a
distribution, and we appropriately construct a geometric graph, the spectrum of
the Laplacian converges to that of the Laplace-Beltrami operator on the
data-manifold. However, relating the partition obtained using the
Laplace-Beltrami operator to the more intuitively defined high-density
clusters can be challenging in general.

Perhaps most similar to our results are the works
\citet{vempala2004,shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain parametric and
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general ensure recovery of density
clusters, which is the focus in our work.

\section{Cluster Quality Critera Bounds for Density Clusters}  
\label{sec: cluster_quality_bounds}

\subsection{Geometric Conditions on Density Clusters}

In order to provide meaningful bounds on the normalized cut and inverse mixing
time of an empirical density cluster, we must introduce conditions on the
density $f$. Let $B(x,r) = \{y \in \Rd: \norm{y - x} \leq r\}$ be the closed
ball of radius $r \geq 0$, centered at $x \in \Rd$.  Given a set $\Aset
\subseteq \Rd$ and $\sigma > 0$, define $\Asig = \Aset + B(0,\sigma) = \{y \in
\Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma\}$, which we call the
$\sigma$-expansion of $\Aset$.

We are now ready to give the conditions, stated with respect to a density
cluster $\Cset \in \Cbb_f(\lambda)$, for some threshold $\lambda > 0$, and an
expansion parameter $\sigma > 0$.

\begin{enumerate}[label=(A\arabic*)]
\item
\label{asmp: bounded_density}
\emph{Bounded density within cluster:} There are $0 < \lambda_{\sigma} <
\Lambda_{\sigma} < \infty$ such that
$$
\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq
\Lambda_{\sigma}.
$$
% and 
% \begin{equation*}
% \frac{\diam \Asig}{\sigma} \leq \mu
% \end{equation*}
% where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

\item 
\label{asmp: low_noise_density}
\emph{Low noise density:} There exist $\gamma, c_1 > 0$ such that for all $x
\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
$$
  \inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \dist(x, \Csig)^{\gamma},
$$
where \smash{$\dist(x,\Aset) = \inf_{x_0 \in \Aset} \norm{x - x_0}$}.
% RJT: should we change this to c_0?  c_1 seems to be used below and throughout
% for other constants, and c_0 is not used

\item
\label{asmp: cluster_separation}
\emph{Cluster separation:}
For all $\Cset' \in \Cbb_f(\lambda)$,
$$
\dist(\Csig,\Csig') > \sigma,
$$
where \smash{$\dist(\Aset,\Aset') = \inf_{x \in \Aset} \dist(x,\Aset')$}.

\item
\label{asmp: cluster_diameter}
\emph{Cluster diameter:}
There exists $D < \infty$ such that for all $x, x' \in \Csig$,
$$
\norm{x - x'} \leq D.
$$
\end{enumerate}

Note that $\sigma$ plays several roles here, precluding arbitrarily narrow
clusters and long clusters in \ref{asmp: bounded_density} and \ref{asmp:
  cluster_diameter}, flat densities around the level set in \ref{asmp: 
  low_noise_density}, and poorly separated clusters in \ref{asmp:
  cluster_separation}.

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} and \ref{asmp:  
cluster_separation} combined are similar to the $(\sigma,\epsilon)$-saliency of
\citet{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citet{singh2009} (as well as
many other works on density clustering and level set estimation.) It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as PPR.

In the next several subsections, we will derive bounds on the cluster quality
criteria evaluated on (thickened) density clusters. For notational simplicity,
hereafter for $S \subseteq V$, we will abbreviate $\Phi(S; G_{n,r})$ by
$\Phi_{n,r}(S)$, and similarly, $\Psi(S; G_{n,r})$ by $\Psi_{n,r}(S)$, and
$\tau_{\infty}(G_{n,r}[S])$ by $\tau_{n,r}(S)$. We will also use $\nu$ for
Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ for the measure of the unit ball
$B=B(0,1)$.  
% RJT: you had B(0,d) before ... just check this should have been B(0,1)

\subsection{Upper Bound on Normalized Cut}

We start with an upper bound on the normalized cut \eqref{eqn: norm_cut} of 
$\Cset_\sigma[\Xbf]$. (In the theorem, the upper bound on the density in
Assumption \ref{asmp: bounded_density} will not actually be needed, so we omit
the parameter $\Lambda_\sigma>0$ from the theorem statement.)

\begin{theorem}
\label{thm: conductance_upper_bound}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation}, for some 
$\sigma, \lambda_{\sigma}, c_1, \gamma > 0$. Then for any $0 < r <
\sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and
\begin{equation}
\label{eqn: conductance_sample_complexity}
n \geq \frac{9\log(2/\delta)}{\epsilon^2}\left(\frac{1}
  { \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
\end{equation}
we have
\begin{equation}
  \label{eqn: conductance_additive_error_bound}
  \frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq 4 c_{\sigma} d
  \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
    \frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
\end{equation}
with probability at least $1-\delta$, where $c_{\sigma} = 1 / \sigma$. 
\end{theorem}

\begin{remark}
The proof of Theorem \ref{thm: conductance_upper_bound}, along with all other
proofs in this paper, can be found in the supplementary document. The key 
idea is that for any $x \in \Cset$, the simple (possibly loose) fact
$B(x,\sigma) \subseteq \Csig$ translates into the upper bound $\nu(\Cset_{\sigma
  + r}) \leq (1 + 2d r /\sigma)\nu(\Csig)$. We leverage \ref{asmp:
  low_noise_density} to find a corresponding bound on the weighted volume,
then apply standard concentration inequalities to convert from population-
to sample-based results.  
\end{remark}

\begin{remark}
The inequality in \eqref{eqn: conductance_additive_error_bound} is almost
tight. Specifically, choosing $\Asig = B(0,\sigma)$ and
$$
f(x) = \lambda - \dist(x,\Asig)^{\gamma},
$$
% RJT: is this correct (is this what you meant)?  Original formula is below
% $$
% f(x)  = \begin{cases}
%   \lambda &\text{for $x \in \Asig$}, \\
%   \lambda - \dist(x,\Asig)^{\gamma} & \text{ for $0 < \dist(x,\Asig) < r$}, 
% \end{cases}
% $$
we have that for $n$ on the order of the lower bound in \eqref{eqn: 
  conductance_sample_complexity}, 
$$
  \frac{\Phi_{n,r}(\Asig[\mathbf{X}])}{r} \geq c \frac{(\lambda -
    \frac{r^{\epsilon+1}}{\epsilon+1})}{\lambda} - \epsilon,
$$
with probability at least $1 - \delta$, for some constant $c$. (Note that the
factor of $1 / \sigma$ in $c_{\sigma}$ is not replicated above.)
\end{remark}

\subsection{Lower Bound on Inverse Mixing Time}

Next we lower bound the inverse mixing time \eqref{eqn: inv_mixing_time} of 
$\Csig[\Xbf]$, or equivalently, as $\Psi_{n,r}(\Csig[\Xbf]) =
1/\tau_{n,r}(\Csig[\Xbf])$, we upper bound the mixing time. 
% RJT: note I changed the next two theorems to be upper bounds on the mixing
% time.  Upper bounds on the inverse inverse mixing time, which was a bit
% superfluous.  Just make sure that the supplement is in sync with this

\begin{theorem}
\label{thm: inverse_mixing_time_lower_bound_nonconvex}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D > 0$. Then for any $0 <
r < \sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and $n$ satisfying 
$$
\sqrt{3^{d+1}\frac{(\log n + d\log \mu + \log(4/\delta))}{n \nu_d r^d
    \lambda_{\sigma}}} \leq \epsilon, 
$$
where \smash{$\mu = \log(\frac{2D}{r})$}, we have
\begin{multline}
\label{eqn: inverse_mixing_time_lower_bound_nonconvex}
\tau_{n,r}(\Csig[\Xbf]) \leq (d \log \mu + c_{\lambda}) \cdot \\  
\left(c_1 + c_2 \frac{\Lambda_{\sigma}^4D^{2d}}{\lambda_{\sigma}^4r^{2d}}\left(d
    \log \mu + c_{\lambda}\right)\right),
\end{multline}
with probability at least $1 - \delta$, where $c_1,c_2>0$ are constants
depending only on the dimension $d$, and $c_{\lambda} = \log(\Lambda_{\sigma}^2/ 
\lambda_{\sigma}^2)$.   
\end{theorem}
% RJT: saying that c_1,c_2 depend on the dimension is a little odd/weak.
% c_2 multiplyies, e.g., a factor of d \log \mu.  So if it's d^{10}, then this
% will swamp the linear factor of ....

\begin{remark}
The proof of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} relies
on upper bounding the mixing time using the \emph{conductance} of $G_{n,r}[\Csig[X]]$, 
$$
\widetilde{\Phi} = \min_{S \subseteq \Csig[\Xbf]} \Phi(S; G_{n,r}[\Csig[X]]).
$$
The factor of $1/r^{2d}$ in the bound in \eqref{eqn:
  inverse_mixing_time_lower_bound_nonconvex} is suboptimal. This exponential 
dependence on $d$ stems from a loose bound on the aforementioned
conductance. In particular, we assert only that any set $S \subseteq \Csig[X]$
must have $\cut(S; \Csig[\Xbf])$ on the order of $n^2 r^{2d}$, while upper bounding
$\vol(S; \Csig[\Xbf])$ by roughly $n^2 r^d$, for a bound on the conductance of
order $r^d$. The presence of $r^{2d}$ comes from upper bounding the mixing time
by about \smash{$1/\widetilde{\Phi}^2$}, this being a variant of classic results
on rapid mixing \citep{jerrum89}.  
\end{remark}

\subsection{Tighter Lower Bound Under Convexity}

It is possible to sharpen the dependency on $d$ in 
\eqref{eqn: inverse_mixing_time_lower_bound_nonconvex}, but at the cost of an  
additional assumption.
\begin{enumerate}[label=(A\arabic*)]
\setcounter{enumi}{4}
\item 
\label{asmp: convex}
\emph{Convexity:} The set $\Cset_\sigma$ is convex. 
\end{enumerate}

With \ref{asmp: convex} in place, we give a tighter bound on the
mixing time (albeit one that holds only asymptotically).

\begin{theorem}
\label{thm: inverse_mixing_time_lower_bound}
Assume the conditions of Theorem \ref{thm:
  inverse_mixing_time_lower_bound_nonconvex}, and additionally, \ref{asmp:
  convex}. Then for any $0 < r < \sigma/4d$, the following holds with
probability 1:   
\begin{align}
\label{eqn: inverse_mixing_time_lower_bound}
  \limsup_{n \to \infty} \, &\tau_{n,r}(\Csig[\Xbf]) \leq \nonumber \\ 
&\quad c_1 \frac{\Lambda_{\sigma}^8}{\lambda_{\sigma}^8} (\log \mu +
  c_{\lambda}) \biggl( c_{\lambda}(d^3 \log\mu + c_2) +{} \nonumber \\  
&\quad \frac{d^2D^2}{r^2} \bigl( \log d + c_2
  \frac{\Lambda_{\sigma}^{2}}{\lambda_{\sigma}^{2}}c_{\lambda} +{} \nonumber \\  
&\quad c_3  + \log \log \mu \bigr) \biggr) + c_d \frac{o_r(1)}{r^2},
\end{align}
where \smash{$\mu = \log(\frac{2D}{r})$}, $c_1,c_2,c_3>0$ are universal
constants, $c_d$ is constant in $r$ but depends on dimension $d$, and
$c_{\lambda} = \log(\Lambda_{\sigma}^2/ \lambda_{\sigma}^2)$.  
% Additionally, $\lim_{r \to 0} o_r(1) = 0$.  
\end{theorem}

\begin{remark}
The only potentially exponential dependence on dimension comes from the
factor of $c_d$. However, for sufficiently small values of $r$ this will
be dominated by the preceding factors, due to the presence of the
$o_r(1)$ term. 
\end{remark}
% RJT: Don't get this comment.  You have o_r(1) / r^2, so this might still blow
% up as r -> 0, right?

>>>>

\theoremstyle{aldenrmrk}
\newtheorem{remark}{Remark}

\newcommand{\theHalgorithm}{\arabic{algorithm}}

\icmltitlerunning{Local clustering of density upper level sets}

@@

\twocolumn[
\icmltitle{Local Spectral Clustering of Density Upper Level Sets}

\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Alden Green}{cmu}
\icmlauthor{Sivaraman Balakrishnan}{cmu}
\icmlauthor{Ryan Tibshirani}{cmu}
\end{icmlauthorlist}

\icmlaffiliation{cmu}{Department of Statistics and Data Science, Carnegie Mellon University, Pittsburgh PA, USA}

\icmlcorrespondingauthor{Alden Green}{ajgreen@andrew.cmu.edu}

\icmlkeywords{local clustering}

@@

\section{Empirical Performance of PPR on Gaussian Mixture Models}

The assumptions and theory of Section \textcolor{red}{2} are tailored towards density functions with sharp transitions in gradient around the perimeter of the density cluster. These type of functions will satisfy \textcolor{red}{A3} with a small ratio of $\Lambda_{\sigma}$ to $\lambda_{\sigma}$, while still satisfying \textcolor{red}{A4}, potentially for $\gamma << 1$.

As mentioned previously, one line of work on spectral algorithms assesses their performance on mixture models. Particularly well-developed is the characterization of clustering performance on Gaussian Mixture Models (GMMs). Gaussian Mixture Models have smooth derivatives, and are therefore not good candidates for our work (at least, for reasonable values of the parameters in \textcolor{red}{(A1)-(A4)} ). Since they are a classical choice in the clustering literature, we investigate \pprspace performance on them empirically.

\begin{figure*}[ht]
	\vskip 0.2in
	\begin{center}
	\centerline{\includegraphics[width=\columnwidth]{graphic_1}}
	\caption{Algorithm \ref{alg: ppr} run over 4 example datasets. Parameters $\alpha$ and $\pi_0$ were tuned to minimize normalized cut of output set. Colors represent \pprspace cuts; shapes correspond to the maximum density among mixture components. Sample size $n = 400$.}
	\label{fig: graphic_1}
	\end{center}
	\vskip 0.2in
\end{figure*}

Figure \ref{fig: graphic_1} tells a story with roughly the same narrative as our theoretical work. The left two plots demonstrate instances, in one and two-dimensions, respectively, where the \pprspace cut has small error relative to a density cut. The right two plots, by contrast, show \pprspace failing to recover, even approximately, a density cut.

In the bottom right panel in particular, we see the effect of a high diameter on the performance of \pprspace. Although any one vertex on the boundary of a density level set may have few cut edges, the length of the boundary is sufficiently long to make the entire cut too large for \pprspace to select it. Instead, the \pprspace cut defaults to a more geometrically compact shape, which has much smaller (unweighted) perimeter.
